The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2022-02-14 Manuscript/Response_To_Comments.tex

\begin{comment}
 Why are two pointers singled out in the data? Any insight on why 2P\% is 
 affected more? I would provide some insight. My hypothesis is officiating 
 bias as shown in many papers (less discretion on 3P fouls). 
\end{comment}

 We thank the reviewer for bringing this concern to our attention. We realize
 we missed discussing the significance of studying each type of shot individually,
 not just 2-pointers. Discussion on the importance of examining each of the three
 types of shot was added, including the 2-point shot, at the end of the first 
 paragraph in the Data section.

\begin{comment}
Comment a bit on these effect sizes. Does a 3\% decrease in 2P\% matter in a game?
That may only be 1 shot. While it's statistically significant, I wonder 
if it's practically significant.
\end{comment}

 A paragraph was added to the Discussion section (fourth paragraph) detailing the practical 
 significance of the 3\% change in 2-point shooting. In short, many NBA games 
 are decided by 6 points or less and in those games late-game strategy is 
 completely dependent exactly how many points one team is behind. Even a small
 shift in score can drastically shift probability of outcomes and strategy 
 for teams. The paper now goes into detail on how a 3-point swing in these 
 close games can completely change the game.

@@

 We thank the reviewer for being appreciative and understanding of the 
 good motivations driving this paper. We agree that a deeper discussion 
 of why home-court advantage would not exist in the bubble would be beneficial 
 to setting up the paper. We added in a discussion of why all the different 
 factors play into creating home-court advantage and how each one may or may 
 not still exist in the bubble. This addition to the paper can be found in the first paragraph
 of the Introduction section. For the most part, these factors disappeared 
 in the context of the bubble, except travel/time and miles for reasons we
 discuss in the paper.

\begin{comment}
Also, this paper https://arxiv.org/abs/1701.05976 finds that altitude matters
(Colorado and Utah teams have the highest home advantage in multiple sports),
so that could be added/discussed. So changes in home advantage in the bubble
could impact teams differently, and NBA teams Nuggets and Jazz could be hurt
more than other teams since they no longer benefit from an altitude advantage.
\end{comment}

We are very grateful that referee~2 brought \citet{Lopez} to our attention. 
It was added to Literature Review (at the end of the third paragraph) and
discussed in the Introduction(first paragraph) when setting 
up and explaining the factors that would be missing from normal home-court 
advantage when playing in the NBA bubble. The most noteworthy point added 
from this paper, as pointed out by referee~2, was the evidence supporting 
geographic factors impact home-court advantage. Particularly, teams like 
the Utah Jazz and Denver Nuggets are the NBA teams who would benefit from 
the high-altitude effects. This is discussed in more detail in the manuscript.

In addition, a few other references were added to the Literature Review section.
First, \cite{Price} discuss empirical evidence in favor of referee bias towards
home teams. Second, \cite{Jones} analyzes the effects of home-court advantage
as component parts and how the effects of home-court advantage impact each 
individual quarter of play. Lastly, \cite{Boudreaux} uses a natural
experiment to estimate how the crowd effect of home-court advantage impacts winning.

@@

\begin{comment}
In your plots with histograms, it would be useful to have a vertical 
line showing the mean to visually compare
the two distributions.

It seems like you made this using \texttt{facet\_grid(shot.type ~ season)} or
something similar. I think it would be easy to add vertical lines fairly 
easily by using \texttt{group\_by} and summarise to make a data frame df.means
like this

@@

2017-2019 2p xx.x

2017-2019 3p xx.x

2017-2019 ft xx.x

and then adding
\texttt{geom\_vline(data=df.means, aes(x=mean.perc)) +}
somewhere in your ggplot code.

@@

If the columns season and shot type are the same as in your other data
set, and the same names used in facet\_grid, a different vertical line
should show up for each season and shot type.
\end{comment}

 While this was not the exact coding solution used to add reference lines
 to the histograms, it was greatly appreciated and was the perfect stepping 
 stone to finding the code solution. As mentioned earlier, the code was
 added to include a reference line at the mean on each histogram we provided,
 within the Data section of the manuscript.

\begin{comment}
If you use a regression, you could take into account strength of team,
strength of opponent, and home advantage simultaneously. This could 
be done with any of the metrics you considered. For example,
\texttt{Score \texttildelow 1 + HomeTeamOnOffenseIndicator + OffensiveTeam + DefensiveTeam}.
Did you consider a regression? Why would it, or would it not, be useful? 
If strength of opponent (in the example above, the DefensiveTeam) doesn’t
vary much from team to team (in other words, if there is a
balanced schedule), then it might not be that helpful to account for 
strength of opponent. If it does vary a lot, then it might help.
In a typical regular season, the NBA schedule is pretty balanced. 
Is the schedule a balance schedule in the bubble?
Would accounting for team strength change the home advantage estimates? If so, 
in what situations would it increase home advantage estimates? Decrease?
You don’t necessarily have to do all of this analysis, but you should at least 
discuss somewhere why you didn’t use a regression, and justify your decision.
I’m not saying that a regression is necessarily better, and the answers to 
the above questions might suggest it wouldn’t help. But a lot of people
would think of estimating home advantage using a regression by including a home 
indicator like the example above, and then using the hypothesis test where 
the null hypothesis is that the regression coefficient for home indicator
is zero. So it is important to, at minimum, discuss your choice to not use
a regression, and justify that
choice.
\end{comment}

### 2022-02-14 Manuscript/NBABubbleAnalysis.Rmd
final check

(ref:cap-Fig1) Winning percentage of NBA home teams in the playoffs since 2010, the green line denotes .500
```{r Fig1, echo=FALSE, fig.cap = " (ref:cap-Fig1)", fig.align='center'}
winyear <- cbind(winpct, years)
winyear <- as.data.frame(winyear)
highlight <- data.frame(winpct = .482, years = 2020)
pdf("Fig1.pdf")
ggplot(winyear, aes(x=years, y=winpct))+
      geom_point()+
      geom_point(data=highlight,
             aes(x=years, y=winpct),
             color='red')+
      scale_x_continuous(name = "Year", breaks = years, labels = years)+
      scale_y_continuous(name =  "Winning Percentage", limits = c(0.4, 0.8))+
      geom_hline(aes(yintercept = 0.5), color = "green")+
      geom_text(aes(label=winpct), hjust=0, vjust=-1, size=2)+
      theme(aspect.ratio=3/4)
```

@@

pdf("Fig2.pdf")
ggplot(combined) +
  geom_histogram(aes(x = PTS, y = ..density.., fill = Home.Away), bins = 10, binwidth = 5)+
  facet_grid(Year ~ Home.Away)+
  theme(legend.position = "none")+
  geom_vline(data=ref_line2, aes(xintercept=PTS))+
  geom_text(data=ref_line2, aes(x=PTS+3.5, label="mean", y=0.045))+
  theme(aspect.ratio=3/4)
```

@@

(ref:cap-Fig3) Histograms of home shooting percentages for two-
pointers, three-pointers and free throws for 2020 (top) vs. 2017-19 (bottom)
```{r Fig3, echo=FALSE, fig.cap="(ref:cap-Fig3)"}
hometwo2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_home$X2P.)
homethree2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_home$X3P.)
homeFT2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_home$FT.)
hometwo2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_home$X2P.)
homethree2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_home$X3P.)
homeFT2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_home$FT.)
homegraph<-rbind(hometwo2020, homethree2020, homeFT2020, hometwo2017_2019, homethree2017_2019,
                 homeFT2017_2019)
ref_line3 <- aggregate(data=homegraph, Percentage ~ Year + Type, FUN=mean)
pdf("Fig3.pdf")
ggplot(homegraph) +
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#00BFC4", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)+
  geom_vline(data=ref_line3, aes(xintercept=Percentage))+
  geom_text(data=ref_line3, aes(x=Percentage+6, label="mean", y=0.075))+
  theme(aspect.ratio=3/4)
```

(ref:cap-Fig4) Histograms of away shooting percentages for two-
pointers, three-pointers and free throws for 2020 (top) vs. 2017-19 (bottom)"
```{r Fig4, echo=FALSE, fig.cap="(ref:cap-Fig4)", fig.align='center'}
awaytwo2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_away$X2P.)
awaythree2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_away$X3P.)
awayFT2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_away$FT.)
awaytwo2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_away$X2P.)
awaythree2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_away$X3P.)
awayFT2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_away$FT.)
awaygraph<-rbind(awaytwo2020, awaythree2020, awayFT2020, awaytwo2017_2019, awaythree2017_2019,
                 awayFT2017_2019)
ref_line4 <- aggregate(data=awaygraph, Percentage ~ Year + Type, FUN=mean)
pdf("Fig4.pdf")
ggplot(awaygraph) +
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#F8766D", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)+
  geom_vline(data=ref_line4, aes(xintercept=Percentage))+
  geom_text(data=ref_line4, aes(x=Percentage+6, label="mean", y=0.075))+
  theme(aspect.ratio=3/4)
```

>>>>

(ref:cap-Fig1) Winning percentage of NBA home teams in the playoffs since 2010, the green line denotes .500
```{r Fig1, echo=FALSE, fig.cap = " (ref:cap-Fig1)", fig.align='center'}
winyear <- cbind(winpct, years)
winyear <- as.data.frame(winyear)
highlight <- data.frame(winpct = .482, years = 2020)
# pdf("Fig1.pdf")
p <- ggplot(winyear, aes(x=years, y=winpct))+
    geom_point()+
    geom_point(data=highlight,
               aes(x=years, y=winpct),
               color='red')+
    scale_x_continuous(name = "Year", breaks = years, labels = years)+
    scale_y_continuous(name =  "Winning Percentage", limits = c(0.4, 0.8))+
    geom_hline(aes(yintercept = 0.5), color = "green")+
    geom_text(aes(label=winpct), hjust=0, vjust=-1, size=2)
#    theme(aspect.ratio=3/4)
ggsave("Fig1.pdf", plot = p, height = 4.2, width = 6)
```

@@

# pdf("Fig2.pdf")
p <- ggplot(combined) +
  geom_histogram(aes(x = PTS, y = ..density.., fill = Home.Away), bins = 10, binwidth = 5)+
  facet_grid(Year ~ Home.Away)+
  theme(legend.position = "none")+
  geom_vline(data=ref_line2, aes(xintercept=PTS))+
    geom_text(data=ref_line2, aes(x=PTS+3.5, label="mean", y=0.045))
ggsave("Fig2.pdf", plot = p, height=4.2, width = 6)
```

@@

(ref:cap-Fig3) Histograms of home shooting percentages for two-
pointers, three-pointers and free throws for 2020 (top) vs. 2017-19 (bottom)
```{r Fig3, echo=FALSE, fig.cap="(ref:cap-Fig3)"}
hometwo2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_home$X2P.)
homethree2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_home$X3P.)
homeFT2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_home$FT.)
hometwo2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_home$X2P.)
homethree2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_home$X3P.)
homeFT2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_home$FT.)
homegraph<-rbind(hometwo2020, homethree2020, homeFT2020, hometwo2017_2019, homethree2017_2019,
                 homeFT2017_2019)
ref_line3 <- aggregate(data=homegraph, Percentage ~ Year + Type, FUN=mean)
# pdf("Fig3.pdf")
p <- ggplot(homegraph) +
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#00BFC4", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)+
  geom_vline(data=ref_line3, aes(xintercept=Percentage))+
  geom_text(data=ref_line3, aes(x=Percentage+6, label="mean", y=0.075))
ggsave("Fig3.pdf", plot = p, height=4.2, width = 6)
```

(ref:cap-Fig4) Histograms of away shooting percentages for two-
pointers, three-pointers and free throws for 2020 (top) vs. 2017-19 (bottom)"
```{r Fig4, echo=FALSE, fig.cap="(ref:cap-Fig4)", fig.align='center'}
awaytwo2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_away$X2P.)
awaythree2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_away$X3P.)
awayFT2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_away$FT.)
awaytwo2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_away$X2P.)
awaythree2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_away$X3P.)
awayFT2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_away$FT.)
awaygraph<-rbind(awaytwo2020, awaythree2020, awayFT2020, awaytwo2017_2019, awaythree2017_2019,
                 awayFT2017_2019)
ref_line4 <- aggregate(data=awaygraph, Percentage ~ Year + Type, FUN=mean)
# pdf("Fig4.pdf")
p <- ggplot(awaygraph) +
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#F8766D", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)+
  geom_vline(data=ref_line4, aes(xintercept=Percentage))+
    geom_text(data=ref_line4, aes(x=Percentage+6, label="mean", y=0.075))
ggsave("Fig4.pdf", width=6, height=4.2)
```
